Effectiveness of Instant Feedback on Developer Navigation Practices

### 2013-09-02 paper_id1.tex
Roughly 90% complete on assessment

\section{Introduction}
%update
Deploying these techniques requires communicating and motivating developers to use them.  Murphy-Hill and colleagues  \cite{wbsnipes:Hill2011Peer} show that developers can effectively learn about new tools from their peers by watching them work and through recommendations, yet this happens very rarely, in part because developers rarely work on the same computer at the same time.  This personal connection limits the knowledge transfer to the co-located circle of colleagues connected with the early adopters.  
%update
Finding a way to automate the communication of technique recommendations between developers enhances deployment of best practices in a globally distributed software development organization.   
%PF the following sounds redundant
To best match the peer recommendation scenario described above, the method should mimic the personal recommendation model while rewarding successful adopters.

@@

\subsection{Assessment Background}
When discussing gamification, folks raise a concern about replacing an activity that is intrinsically movitated with extrinsic motivation provided by points and achievements.  To discover what is behind this concern, several studies are useful to understand what should be done.

ii.	Determine through a questionnaire survey how developers are motivated to use training, practices, and tools and whether they are open to having that information shared with their peers and managers with recognition by badges and points. The developer study conducted with over 130 software developers at ABB challenged whether achievement motivation would help software engineers and whether they would accept sharing data on tools and practices they use to help others.

@@

Using a question from the Saachi and Saachi survey\cite{wbsnipes:SaatchiGameification} as directly as possible, we asked developers at ABB "How interested they would be in working for a company that incorporated some aspects of games into software engineering tools as a way to increase productivity in the workplace"?  We segmented resuults by country and found that gamification of software engineering is more interesting to developers in India, Poland, and Finland than in other major countries.  The responses indicating that developers were at least somewhat interested is 75-88\% In India, Poland, and Finland.  For the US, Sweden, and Switzerland the responses indicate 60-65\% are at least somewhat interested.  The overall ABB response is comparable to the Saachi and Saaschi survey\cite{wbsnipes:SaatchiGameification} upon which this question was based that found 75\% of 18-45 year olds were at least somewhat interested.

95\% of the developers surveyed are likely to try tools and practices recommended to them through an automated usage tracking system.  This indicates that a recommender system would positively impact the deployment of software engineering tools and practices at ABB.

One area we questioned was whether sharing this information would be a concern for developers.  A graduated scale was used for the community that would be shared with and distinction between whether that sharing was anonymous or the information was personally identifiable.  Over 90\% of respondents are comfortable with sharing anonymous/non-anonymous information with their team members.  The figure below shows how this comfort level decreases as the scope of who the information is shared with increases.  People were less comfortable sharing with anyone (the category for people outside ABB) particularly if the data were not anonymous.  
Two conclusions from this feedback are that people are very willing to share information that could help their team. Second, sharing within the company is acceptable if we take care to make the data anonymous.

Developers were asked what would motivate them to use tools and practices and given 4 choices to rank.  The top motivator for using tools and practices is coworker recommendation with 75\% of respondents ranking in 1st or 2nd place.   The lowest was having badges posted on their You@ABB profile where 33\% ranked this in 1st or 2nd place.   Badges also ranked low in motivations for taking more software engineering training.  We conclude that a facility to share coworker recommendations on tools and practices has the best opportunity to increase their adoption.  Since a competition among teams ranks slightly higher than management mandate, we have the choice to apply one or the other or both as motivation for using key practices.

@@

Based on the survey results we draw the following conclusions for this project going forward:
•	95\% of developer responding would try tools and practices suggested by an automated recommender system particularly if they were provided in context with the current workflow pattern the developer follows.
•	Responses indicate a preference for team competition over individual awards (badges) and awards for quality practices ranked highest.  So, the tool will incorporate team competition ideas and recognize quality practices.
•	Over 90\% of respondents approved of sharing data with their team members

@@

Gamification motivates people to achieve goals through incremental scoring and discrete awards that are typically shared with peers in an on line community.  Companies are gameifying their business processes and even public facing web sites to encourage people to complete a task such as their profile, stay loyal by earning rewards (air miles), or participate frequently by being part of a community.  Aspects of Gamification that should be considered when applying this popular technique to a process or system include motivational alignment, whether to include the competitive dimension, and making achievements encouraging.

First it is important to consider the existing motivation people may have for performing the activity.  Applying gamefiication to activities that people are motivated intrinsically to do may decrease their motivation.  For example applying gamification to education is very difficult because learners are typically intrinsically motivated more than extrinsically motivated to achieve in their education \cite{wbsnipes:Vansteenkiste2008Does}.    In other words they are mostly interested in learning to better themselves rather than earn the highest GPA.   If an activity is primarily extrinsically motivated then gamification is an easy fit to directly apply to the activity.

@@

Using these guidelines we assigned points to structured navigation commands which received 1-10 points each while unstructured navigation commands received 0 points.  Assigning negative points while possible, is discouraged as this may demotivate folks who do not have detailed knowledge of how the study is designed.  When assigning points, make sure you recognize the relationship between points and levels such that "leveling up" happens at a frequent enough pace to make that motivating for users during the study.  Ideally users would level up after a working day or two worth of activity initially extending to a week or even a month following an exponential curve \cite{wbsnipes:Hamari2011Framework} making the game challenging for beginners and more experienced developers.  Designing the points and levels such that above average developers can complete all levels during the study period is ideal.

@@

After selecting the practice to focus on, we needed to select an intervention method.  The term intervention, indicates that we are seeking to influence developers to perform a practice previously defined by software engineering community.  Intervention involves communication about the practice in a way that motivates adoption \cite{}.  Interventions can include presentations, one-on-one discussions, distance education mechanisms such as webinars, and word of mouth recommendations.  Part of the research objective is to determine whether we can acheive adoption of a practice with alternative interventions that are more deliberate than word of mouth and perhaps more scalable than webinars or presentations.  Utilizing the automation capabilities of Blaze, we chose to automatically pop-up a window with training information on structured navigation practices and tools. For this study we chose to make a nice looking graphic that users click on to see information about navigation commands, obtain and install the Sando search tool, and learn about the Blaze game.  The site uses a combination of video and blog-post style information easily accessible to the developer.  The usefulness of the communication site is rated as part of our study completion survey.

The study established three time periods, a prelinary period to establish a baseline, a period where information is communicated but no immediate feedback or points are displayed, and a period where points are displayed and all information is available.  We chose to deploy the Blaze tool without communicating the study purpose for the baseline period so that an accurate picture of the baseline could be captured by the tool.  Then the tool would pop-up a display with the information on the communication web site asking the developers to consume the information and begin using the practices.   Finally the display of points fo accumulated by the developer provides the instant feedback mechanism.

@@

Johnson and Kou defined Zorro\cite{V:Johnson2007Automated}, a system for detecting whether developers use Test Driven Development techniques based on data from Hackystat.  Zorro divides development activities into episodes delimited by events such as configuration management code check-in, start of a unit test run, or start of a build.  Using the distinct events that developers follow within these episodes, Zorro determines whether the episode followed Test-Driven Development practices per their classic definition of test first - code - test pass, or a different scenario.  In two student-based studies comparing Zorro classifications with a simultaneous observational screen video, Zorro achieved between 70\% \cite{Kou2010Operational} and 89\% \cite{V:Johnson2007Automated} accuracy in classifying episodes into their proper TDD scenarios.  The studies did not, however, attempt to use or evaluate the influence of instant feedback aimed at encouraging participants to follow the classic definition of Test Driven Development.  In our study, we attempt this leap to present data collected directly to developers attempting to influcence them to change their practices.

@@

The Pro Metrics (PROM) tool provides a framework for collecting data for further analysis from tools used by developers.\cite{Coman2009Casestudy}  It provides a flexible data model and a plug-in architecture to facilitate collection from different data sources.  Studies conducted using prom include a series of studies by it's creators on trends in time spent Pair Programming \cite{Coman2008Investigating}, benefits of refactoring on productivity \cite{Moser2008Case}, impact of refactoring on re-usability \cite{Moser2006Does}, and prediction of effort \cite{Abrahamsson2007Effort}.  The PROM tool and studies that use it track the time spent editing files and methods from the IDE and code metrics obtained trough source code analysis.  Applying these data to specific research questions such as whether refectoring improves productivity \cite{Moser2008Case} shows the utility of combining automated effort data with code metric data.  Taking a different focus on techniques, our study used more detailed events captured from the IDE  to detect how the user was navigating through the code in addition to knowing how long the editor is open for a particular file.

Murphy-Hill et. al. study a large usage history data set and apply several different algorithms to accurately suggest to users new commands in their environment to use.\cite{MurphyHill2012Improving} Algorithms based on command history performed well particularly when synchronized chronologically with the recipient's usage history. Novice users were better served by recommender algorithms that did not require a long history, where more expert users benefited from more sophisticated algorithms that included more history. The "Most Widely Used" algorithm that recommends commands based on the collective usage profile of the team performed nearly as well as the more sophisticated algorithms. The paper presents the methodology for making recommendations and the evaluation of the accuracy of the recommendations. Our work goes the next step of evaluating whether a specific recommended practice gets adopted by developers when influenced through gamification methods.

>>>>

\section{Introduction}
%update
Deploying these techniques requires communicating and motivating developers to use them.  Murphy-Hill and colleagues  \cite{wbsnipes:Hill2011Peer} show that developers can effectively learn about new tools from their peers by watching them work and through recommendations, yet this happens very rarely, in part because developers rarely work on the same computer at the same time.  This personal connection limits the knowledge transfer to the co-located circle of colleagues connected with the early adopters.

\subsection{Assessment Background}
When discussing gamification, folks raise a concern about replacing an activity that is intrinsically motivated with extrinsic motivation provided by points and achievements.  To discover what is behind this concern, several studies are useful to understand what should be done.

Maehr proposes an affirmative theory that individuals achieve as a member of a social group choosing the behavior that will meet the expectations and values of the group that is significant to them. \cite{wbsnipes:MaehrCulture}  So the presence of a social group where the person is motivated to belong and excel within is important for creating achievement motivation.  Maehr states that since achievement behaviors can be triggered by circumstance, that they are present in most people regardless of background.

@@

Understanding the concerns with maintaining intrinsic motivation while adding elements of achievement motivation, we developed an assessment survey questionnaire to determine whether the community would be receptive to a gamification approach to software development.   We designed  the survey to assess how developers are motivated to use training, practices, and tools and whether they are open to having that information shared with their peers and managers, and being recognized with badges and points. The developer study conducted with over 130 software developers at ABB challenged whether achievement motivation would help software engineers and whether they would accept sharing data on tools and practices they use to help others.

@@

Using a question from the Saachi and Saachi survey\cite{wbsnipes:SaatchiGameification} as directly as possible, we asked developers at ABB "How interested they would be in working for a company that incorporated some aspects of games into software engineering tools as a way to increase productivity in the workplace"?  We segmented results by country in Figure \ref{fig:gamification} and found that gamification of software engineering is more interesting to developers in India, Poland, and Finland than in other major countries.  The responses indicating that developers were at least somewhat interested is 75-88\% In India, Poland, and Finland.  For the US, Sweden, and Switzerland the responses indicate 60-65\% are at least somewhat interested.  The overall ABB response is comparable to the Saachi and Saaschi survey\cite{wbsnipes:SaatchiGameification} upon which this question was based that found 75\% of 18-45 year olds were at least somewhat interested.

\begin{figure}\begin{mdframed}[linecolor=white]
	\includegraphics[width=3in]{gamificationquestion.pdf}
	\caption{Acceptance of gamification}
	\label{fig:gamification}
\end{mdframed}\end{figure}

We asked developers how likely they would be to try tools and practices recommended to them through an automated usage tracking system.  Answers showed that 95\% of the developers surveyed are likely to try tools and practices recommended to them.  This indicates that a recommendation system would positively impact the deployment of software engineering tools and practices at ABB.

One area we questioned was whether sharing detailed usage information with colleagues and the company as a whole would be a concern for developers.  A graduated scale was used for the community that would be shared with and distinction between whether that sharing was anonymous or the information was personally identifiable. Figure \ref{fig:comfortwithsharing}  shows that over 90\% of respondents are comfortable with sharing either anonymous or non-anonymous information with their team members. 
Their comfort level decreases as the scope of who the information is shared with increases.  People were less comfortable sharing with anyone (the category for people outside ABB) particularly if the data were not anonymous.  
Two conclusions from this feedback are that people are very willing to share information that could help their team. Second, sharing within the company is acceptable if we take care to make the data anonymous.

\begin{figure}\begin{mdframed}[linecolor=white]
	\includegraphics[width=3in]{ComfortWithSharing.pdf}
	\caption{Comfortable with sharing usage information}
	\label{fig:comfortwithsharing}
\end{mdframed}\end{figure}

Developers were asked what would motivate them to use tools and practices and given 4 choices to rank.  The top motivator for using tools and practices is coworker recommendation with 75\% of respondents in Figure \ref{fig:toolandpracticemotivators}   ranking in 1st or 2nd place. The lowest was having badges posted on their social profile where 33\% ranked this in 1st or 2nd place.   Badges also ranked low in motivations for taking more software engineering training.  We conclude that a facility to share coworker recommendations on tools and practices has the best opportunity to increase their adoption.  Since a competition among teams ranks slightly higher than management mandate, we have the choice to apply one or the other or both as motivation for using key practices.

\begin{figure}\begin{mdframed}[linecolor=white]
	\includegraphics[width=3in]{ToolAndPracticeMotivators.pdf}
	\caption{Motivation for using tools and practices}
	\label{fig:toolandpracticemotivators}
\end{mdframed}\end{figure}

@@

Based on the survey results we draw the following conclusions:
•	95\% of developer responding would try tools and practices suggested by an automated recommendation system particularly if they were provided in context with the current workflow pattern the developer follows.
•	Responses indicate a preference for team competition over individual awards (badges) and awards for quality practices ranked highest.  So, the tool will incorporate team competition ideas and recognize quality practices.
•	Over 90\% of respondents approved of sharing data with their team members

@@

Gamification motivates people to achieve goals through incremental scoring and discrete awards that are typically shared with peers in an on line community.  Companies are gamifying their business processes and even public facing web sites to encourage people to complete a task such as their profile, stay loyal by earning rewards (air miles), or participate frequently by being part of a community.  Aspects of Gamification that should be considered when applying this popular technique to a process or system include motivational alignment, whether to include the competitive dimension, and making achievements encouraging.

First it is important to consider the existing motivation people may have for performing the activity.  Applying gamification to activities that people are motivated intrinsically to do may decrease their motivation.  For example applying gamification to education is very difficult because learners are typically intrinsically motivated more than extrinsically motivated to achieve in their education \cite{wbsnipes:Vansteenkiste2008Does}.    In other words they are mostly interested in learning to better themselves rather than earn the highest GPA.   If an activity is primarily extrinsically motivated then gamification is an easy fit to directly apply to the activity.

@@

Using these guidelines we assigned points to structured navigation commands which received 1-10 points each while unstructured navigation commands received 0 points.  Assigning negative points is discouraged as this may demotivate folks who do not have detailed knowledge of how the points are assigned.   In creating levels, following the guidance from Hamari \cite{wbsnipes:Hamari2011Framework}, we built levels following an exponential curve.  At first users level up after a working day or two worth of activity initially extending to a week or even a month following an exponential curve  making the game motivating for beginners and challenging for more experienced developers.  We designed the points and levels such that above average developers can complete all levels during the study period.

@@

After selecting the practice to focus on, we needed to select an intervention method.  The term intervention, indicates that we are seeking to influence developers to perform a practice previously defined by software engineering community.  Intervention involves communication about the practice in a way that motivates adoption \cite{}.  Interventions can include presentations, one-on-one discussions, distance education mechanisms such as webinars, and word of mouth recommendations.  Part of the research objective is to determine whether we can achieve adoption of a practice with alternative interventions that are more deliberate than word of mouth and perhaps more scalable than webinars or presentations.  Utilizing the automation capabilities of Blaze, we chose to automatically pop-up a window with training information on structured navigation practices and tools. For this study we chose to make a nice looking graphic that users click on to see information about navigation commands, obtain and install the Sando search tool, and learn about the Blaze game.  The site uses a combination of video and blog-post style information easily accessible to the developer.  The usefulness of the communication site is rated as part of our study completion survey.

The study established three time periods, a preliminary period to establish a baseline, a period where information is communicated but no immediate feedback or points are displayed, and a period where points are displayed and all information is available.  We chose to deploy the Blaze tool without communicating the study purpose for the baseline period so that an accurate picture of the baseline could be captured by the tool.  Then the tool would pop-up a display with the information on the communication web site asking the developers to consume the information and begin using the practices.   Finally the display of points fo accumulated by the developer provides the instant feedback mechanism.

@@

Johnson and Kou defined Zorro\cite{V:Johnson2007Automated}, a system for detecting whether developers use Test Driven Development techniques based on data from Hackystat.  Zorro divides development activities into episodes delimited by events such as configuration management code check-in, start of a unit test run, or start of a build.  Using the distinct events that developers follow within these episodes, Zorro determines whether the episode followed Test-Driven Development practices per their classic definition of test first - code - test pass, or a different scenario.  In two student-based studies comparing Zorro classifications with a simultaneous observational screen video, Zorro achieved between 70\% \cite{Kou2010Operational} and 89\% \cite{V:Johnson2007Automated} accuracy in classifying episodes into their proper TDD scenarios.  The studies did not, however, attempt to use or evaluate the influence of instant feedback aimed at encouraging participants to follow the classic definition of Test Driven Development.  In our study, we attempt this leap to present data collected directly to developers attempting to influence them to change their practices.

@@

The Pro Metrics (PROM) tool provides a framework for collecting data for further analysis from tools used by developers.\cite{Coman2009Casestudy}  It provides a flexible data model and a plug-in architecture to facilitate collection from different data sources.  Studies conducted using prom include a series of studies by it's creators on trends in time spent Pair Programming \cite{Coman2008Investigating}, benefits of refactoring on productivity \cite{Moser2008Case}, impact of refactoring on re-usability \cite{Moser2006Does}, and prediction of effort \cite{Abrahamsson2007Effort}.  The PROM tool and studies that use it track the time spent editing files and methods from the IDE and code metrics obtained trough source code analysis.  Applying these data to specific research questions such as whether refactoring improves productivity \cite{Moser2008Case} shows the utility of combining automated effort data with code metric data.  Taking a different focus on techniques, our study used more detailed events captured from the IDE  to detect how the user was navigating through the code in addition to knowing how long the editor is open for a particular file.

Murphy-Hill et. al. study a large usage history data set and apply several different algorithms to accurately suggest to users new commands in their environment to use.\cite{MurphyHill2012Improving} Algorithms based on command history performed well particularly when synchronized chronologically with the recipient's usage history. Novice users were better served by recommendation algorithms that did not require a long history, where more expert users benefited from more sophisticated algorithms that included more history. The "Most Widely Used" algorithm that recommends commands based on the collective usage profile of the team performed nearly as well as the more sophisticated algorithms. The paper presents the methodology for making recommendations and the evaluation of the accuracy of the recommendations. Our work goes the next step of evaluating whether a specific recommended practice gets adopted by developers when influenced through gamification methods.
